---
title: "🗺️ Interactive Mapping with mapgl and Shiny"
author: "DS421 – Cartographic Design"
format:
  html:
    code-fold: true
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## 1️⃣ Census Shiny App

#### Load Libraries

```{r}
library(shiny)
library(tidyverse)
library(sf)
library(mapgl)
library(lubridate)
library(bslib)
library(here)
library(tidycensus)
library(ggpubr)
library(plotly)
library(mapview)
library(leaflet)
```

#### View Census Variables

```{r}
acs_vars <- load_variables(year = 2023, dataset = "acs5", cache = TRUE)
```

#### Transform Data

```{r}
# Define raw ACS variable codes
raw_vars <- c(
  median_income = "B19013_001",
  bachelors_count = "B15003_022",  # Count, not percent
  median_gross_rent = "B25064_001",
  total_population = "B01003_001"
)

# Labels for Shiny dropdown (key = column name in wide data)
variable_choices <- c(
  "Median Household Income" = "median_income",
  "Percent with Bachelor's Degree" = "pct_bachelors",
  "Median Gross Rent" = "median_gross_rent"
)

# Pull and transform ACS data
hawaii_data_raw <- get_acs(
  geography = "tract",
  state = "HI",
  variables = raw_vars,
  geometry = TRUE,
  year = 2023
) |>
  filter(GEOID != "15003981200") |>  # Remove Kalawao County
  st_transform(4326)

# Convert to wide format + calculate % with bachelor's
hawaii_data <- hawaii_data_raw |>
  select(GEOID, NAME, variable, estimate, geometry) |>
  st_drop_geometry() |>   # temp drop for pivot
  pivot_wider(names_from = variable, values_from = estimate) |>
  mutate(
    pct_bachelors = 100 * bachelors_count / total_population
  ) |>
  left_join(
    hawaii_data_raw |> select(GEOID, geometry) |> distinct(),
    by = "GEOID"
  ) |>
  st_as_sf()
```

## 🧩 Challenge 1

Can you make a histogram of median rent colored by county. Hint: You will have to make a new column called county based on the values in the NAME column

```{r}
# Create county column from NAME (e.g., "Hawaii County, Hawaii")
hawaii_data <- hawaii_data |>
  mutate(
    county = str_extract(NAME, "^[^,]+")  # text before first comma
  )

ggplot(hawaii_data, aes(x = median_gross_rent, fill = county)) +
  geom_histogram(binwidth = 100, color = "white", alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~ county, ncol = 2) +
  labs(
    title = "Median Gross Rent by County",
    x = "Median Gross Rent (USD)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

```{r}
ggplot(hawaii_data, aes(x = median_gross_rent, fill = county)) +
  geom_histogram(binwidth = 100, color = "white", alpha = 0.7, show.legend = FALSE) +
  labs(
    title = "Median Gross Rent in Hawaii",
    x = "Median Gross Rent (USD)",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

## 🧩 Challenge 2

A. Can you make a scatterplot showing the relationship between median income and median gross rent. Can you use the `stat_cor()`function to show a correlation coefficient and p value

```{r}
library(ggpubr)

p_rent_income <- ggplot(hawaii_data,
                        aes(x = median_income, y = median_gross_rent)) +
  geom_point(alpha = 0.6) +
  stat_cor(      # adds correlation coefficient + p-value
    method = "pearson",   # default; optional, but explicit
    label.x.npc = "left", # position in normalized plot coordinates
    label.y.npc = "top",
    hjust = 0,
    vjust = 1,
    size = 4
  ) +
  labs(
    title = "Median Household Income vs Median Gross Rent (Hawaii Census Tracts)",
    x = "Median Household Income (USD)",
    y = "Median Gross Rent (USD)"
  ) +
  theme_minimal()

p_rent_income
```

```{r}

```

B. Can you make this scatterplot interactive with `ggplotly()`

------------------------------------------------------------------------

\*\* I was unable to make this scatterplot interactive but was researching and found another way to make it interactive without using the ggplotly library \*\*

```{r}
str(hawaii_data)
names(hawaii_data)
```

```{r}
plot_ly(
  data = hawaii_df,
  x = ~median_income,
  y = ~median_gross_rent,
  type = "scatter",
  mode = "markers"
)
```

## 🧩 Challenge 3

Can you make a choropleth map of Hawaii using `mapview()` that shows median rent by census tract

```{r}
mapview(
  hawaii_data,
  zcol = "median_gross_rent",
  layer.name = "Median Gross Rent",
  legend = TRUE,
  alpha.regions = 0.8
)
```

#### Building a Shiny App

We have these 3 variables in hawaii_data, it would be nice if users can select a variable from a dropdown: [Introducing Shiny](https://shiny.posit.co/)

There are many mapping packages we can use with Shiny. The simplest for now is `mapview`. It is good to note that the mapgl package is also designed specifically to [work with advanced shiny apps](https://walker-data.com/mapgl/articles/shiny.html).

```{r}

ui <- fluidPage(
  titlePanel("Hawaii Census Map"),
  sidebarLayout(
    sidebarPanel(
      selectInput("var", "Choose a variable to map:", choices = variable_choices)
    ),
    mainPanel(
      leafletOutput("map", height = "700px")
    )
  )
)

server <- function(input, output, session) {
  output$map <- renderLeaflet({
    mapview(
      hawaii_data,
      zcol = input$var,
      layer.name = names(oahu_choices)[oahu_choices == input$var]
    )@map
  })
}

shinyApp(ui, server)
```

## 🧩 Challenge 4

Can you create a shiny app using your LANDSAT_USA_contiguous_and_Hawaii_7d.csv in the data folder that allows users to filter the map by confidence. For example, the user can select from a side dropdown to only show *high* confidence fires, *medium* confidence fires, or *low* confidence fires. Do not use AI for this and use the example above. AI Bonus: Add a date slider so users can toggle the map by date. Can you add a heatmap layer users can toggle on and off?

```{r}
fires_raw <- readr::read_csv(here("/Users/takaz/OneDrive/DS-421NEW/data/LANDSAT_USA_contiguous_and_Hawaii_7d.csv"))


fires <- fires_raw |>
  mutate(
    acq_date = as.Date(acq_date, format = "%Y%m%d"),
    confidence = factor(confidence, levels = c("low", "medium", "high"))
  ) |>
  filter(!is.na(acq_date), !is.na(latitude), !is.na(longitude))

fires_sf <- st_as_sf(fires, coords = c("longitude", "latitude"), crs = 4326)

ui <- fluidPage(
  titlePanel("LANDSAT Fire Observations (Hawaii & Contiguous US)"),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "confidence",
        "Filter by Confidence:",
        choices = levels(fires$confidence),
        selected = levels(fires$confidence),
        multiple = TRUE
      ),
      sliderInput(
        "date_range",
        "Date Range:",
        min = min(fires$acq_date),
        max = max(fires$acq_date),
        value = c(min(fires$acq_date), max(fires$acq_date)),
        timeFormat = "%Y-%m-%d"
      ),
      checkboxInput("show_heatmap", "Show Heatmap Layer", value = FALSE)
    ),
    mainPanel(
      leafletOutput("fire_map", height = "700px")
    )
  )
)

server <- function(input, output, session) {
  
  filtered_fires <- reactive({
    req(input$confidence, input$date_range)
    f <- fires_sf |>
      filter(
        confidence %in% input$confidence,
        acq_date >= input$date_range[1],
        acq_date <= input$date_range[2]
      )
    f
  })
  
  output$fire_map <- renderLeaflet({
    f <- filtered_fires()
    
    base_map <- leaflet(f) |>
      addProviderTiles("CartoDB.Positron") |>
      setView(lng = -155.5, lat = 20.5, zoom = 6)
    
    #COLOR BY CONFIDENCE
    pal <- colorFactor(
      palette = c("low" = "green", "medium" = "orange", "high" = "red"),
      levels = levels(f$confidence),
      domain = f$confidence
    )
    
    base_map <- base_map |>
      addCircleMarkers(
        color = ~pal(confidence),
        fillColor = ~pal(confidence),
        fillOpacity = 0.7,
        radius = 4,
        popup = ~paste(
          "<b>Confidence:</b>", confidence, "<br>",
          "<b>Date:</b>", as.character(acq_date)
        ),
        group = "Fires"
      )
    
#TOGGLE
    if (input$show_heatmap && nrow(f) > 0) {
      base_map <- base_map |>
        addHeatmap(
          lng = ~as.numeric(st_coordinates(geometry)[,1]),
          lat = ~as.numeric(st_coordinates(geometry)[,2]),
          radius = 8,
          blur = 15,
          opacity = 0.6,
          group = "Heatmap"
        )
    }
    
    base_map |>
      addLayersControl(
        overlayGroups = c("Fires", if (input$show_heatmap) "Heatmap"),
        options = layersControlOptions(collapsed = FALSE)
      )
  })
}

shinyApp(ui, server)
```

```{r}

```

## 🧩 Challenge 5

Can you add the functionality to your census shiny app (new panel) that allows users to pick two census variables to plot on an interactive scatter plot using `ggplotly()` and `stat_cor()`

We should remain careful about our use/potential dependence on AI for coding but for cases like these where we have built all the individual parts, AI can be really helpful in putting them together in a sleek design on a shiny app.

```{r}
ui <- fluidPage(
  titlePanel("Hawaii Census Data Explorer"),
  
  tabsetPanel(
    tabPanel(
      "Map",
      sidebarLayout(
        sidebarPanel(
          selectInput(
            "map_var",
            "Choose a variable to map:",
            choices = variable_choices
          )
        ),
        mainPanel(
          leafletOutput("census_map", height = "700px")
        )
      )
    ),
    
    tabPanel(
      "Scatterplot",
      fluidRow(
        column(
          width = 4,
          selectInput(
            "x_var",
            "X-axis variable:",
            choices = variable_choices
          ),
          selectInput(
            "y_var",
            "Y-axis variable:",
            choices = variable_choices
          ),
          p("The plot shows census tract-level relationships with correlation and p-value.")
        ),
        column(
          width = 8,
          plotlyOutput("scatter_plot", height = "600px")
        )
      )
    )
  )
)

server <- function(input, output, session) {
  

  output$census_map <- renderLeaflet({
    req(input$map_var)
    mapview(
      hawaii_data,
      zcol = input$map_var,
      layer.name = names(variable_choices)[variable_choices == input$map_var],
      legend = TRUE,
      alpha.regions = 0.8
    )@map
  })
  

  output$scatter_plot <- renderPlotly({
    req(input$x_var, input$y_var)
    

    if (input$x_var == input$y_var) {
      return(
        plot_ly() |>
          layout(
            title = "Please select two different variables.",
            xaxis = list(showticklabels = FALSE),
            yaxis = list(showticklabels = FALSE)
          )
      )
    }
    
    p <- ggplot(hawaii_data, aes_string(x = input$x_var, y = input$y_var)) +
      geom_point(alpha = 0.6) +
      stat_cor(
        aes(label = paste(..rr.label.., ..p.label.., sep = "~,~")),
        label.x.npc = "left",
        label.y.npc = "top",
        hjust = 0,
        vjust = 1,
        size = 4
      ) +
      labs(
        title = "Census Tract Relationship",
        x = names(variable_choices)[variable_choices == input$x_var],
        y = names(variable_choices)[variable_choices == input$y_var]
      ) +
      theme_minimal()
    
    ggplotly(p)
  })
}

shinyApp(ui, server)
```
